---
title: "Exploratory analysis of low-pass WGS"
author: 
- Zhan Yinxiu
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 6,
                      fig.height = 6)
```

```{r libraries}
library(caret)
library(ggplot2)
library(stringi)
library(stringr)
library(gridExtra)
```

```{r functions}
# Function to Create Confusion Matrix Plots and Return Grid
create_confusion_matrix_grid <- function(tumor, plasma, samplesid, type="DEL") {
  # Get unique sample IDs
  unique_samples <- unique(samplesid)
  
  # Initialize a list to store plots
  plot_list <- list()
  
  # Loop through each unique sample ID
  for (sample in unique_samples) {
    # Subset data for the current sample
    subset_tumor <- tumor[samplesid == sample]
    subset_plasma <- plasma[samplesid == sample]
    
    # Compute confusion matrix
    cm <- table(Predicted = subset_plasma, Actual = subset_tumor)
    
    # Convert confusion matrix to a data frame for ggplot2
    cm_df <- as.data.frame(as.table(cm))
    
    # Create the heatmap plot
    plot <- ggplot(cm_df, aes(x = Predicted, y = Actual, fill = Freq)) +
      geom_tile() +
      geom_text(aes(label = Freq), color = "white", size = 5) +
      scale_fill_gradient(low = "lightblue", high = "blue") +
      labs(title = paste(type, sample), x = "Plasma", y = "Tumor") +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5))
    
    # Add plot to the list
    plot_list[[sample]] <- plot
  }
  
  # Arrange the plots into a 3x2 grid
  grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 3, nrow = 2))
  
  # Return the grid plot
  return(grid_plot)
}
```

# Objectives

In this report, we aim to evaluate the feasibility and confidence of CNV detection in plasma samples (vesicles) compared to tumor tissue samples using low-pass whole-genome sequencing (WGS). The low-pass WGS data were analyzed with ichorCNA, employing a 1 Mb bin size to estimate tumor fraction and identify copy number variations.

```{r data loading}
wgs_tumor_files = list.files("/hpcnfs/scratch/DIMA/zhan/gpelicci/liquid_biopsy_project/results/cna/", pattern = "tumor.cna.seg", full.names = TRUE) 
wgs_plasma_files = list.files("/hpcnfs/scratch/DIMA/zhan/gpelicci/liquid_biopsy_project/results/cna/", pattern = "plasma.cna.seg", full.names = TRUE) 

tumor_data = NULL
for(file in wgs_tumor_files){
  tmp = read.delim(file)[,c(1,2,3,4,5,6)]
  colnames(tmp) = c("chr", "start", "end", "ncopy", "status", "logR")
  tmp$sample = str_split(basename(file), "-")[[1]][1]
  tumor_data = rbind(tumor_data, tmp)
}

plasma_data = NULL
for(file in wgs_plasma_files){
  tmp = read.delim(file)[,c(1,2,3,4,5,6)]
  colnames(tmp) = c("chr", "start", "end", "ncopy", "status", "logR")
  tmp$sample = str_split(basename(file), "-")[[1]][1]
  plasma_data = rbind(plasma_data, tmp)
}

```


We start by comparing the deletion and amplification detected in tumor and plasma samples. At this stage, we use all the CNVs.

# Deletion

Below, we present a comparison between plasma and tumor WGS low-pass data in detecting deletions, represented using confusion matrices.

A confusion matrix is a summary of classification outcomes, showing how well predictions align with the actual values. It consists of four components:

- True Positives (TP): Cases where both plasma and tumor samples correctly identify the presence of a deletion.
- True Negatives (TN): Cases where both plasma and tumor samples correctly identify the absence of a deletion.
- False Positives (FP): Cases where a deletion is detected in plasma but not in the tumor.
- False Negatives (FN): Cases where a deletion is detected in the tumor but not in the plasma.

In this analysis:

TRUE indicates the presence of a deletion.
FALSE indicates the absence of a deletion.

As evident from the matrices, there is minimal overlap between deletions detected in plasma and those detected in tumor samples. This highlights the challenge of achieving concordant results between these two sample types, particularly in low-pass WGS data.

```{r, fig.height=8, fig.width=12}
tmptumor = tumor_data$ncopy == 1
tmpplasma = plasma_data$ncopy == 1
samplesid = plasma_data$sample
gridplot <- create_confusion_matrix_grid(tmptumor, tmpplasma, samplesid)
```



# Amplification

In this analysis:

TRUE indicates the presence of a amplification higher or equal than 3 copies
FALSE indicates the absence of a amplification.

As evident from the matrices, there is little overlap between amplifications detected in plasma and those detected in tumor samples. This highlights the challenge of achieving concordant results between these two sample types, particularly in low-pass WGS data.

```{r, fig.height=8, fig.width=12}
tmptumor = tumor_data$ncopy >= 3
tmpplasma = plasma_data$ncopy >= 3
samplesid = plasma_data$sample
gridplot <- create_confusion_matrix_grid(tmptumor, tmpplasma, samplesid, "Amplification")
```


## Comparison of log ratio of copy number

Next, we examined how the log ratio of copy number in each bin varies between plasma and tumor samples, stratifying the analysis based on tumor sample categories: gain, neutral, and loss of copy number. At a resolution of 1 Mb, we observed a general tendency for copy number ratios to be preserved between plasma and tumor samples. Regions with copy number gains consistently exhibited higher log ratios compared to neutral or deleted regions. However, given the low tumor fraction (<10%) and sequencing depth, detecting CNAs at a 1 Mb resolution may exceed the limits of reliable detection. To address this, we will repeat the analysis using a coarser resolution of 10 Mb.

```{r, fig.height=8, fig.width=12}

  # Get unique sample IDs
  unique_samples <- unique(samplesid)
  
  # Initialize a list to store plots
  plot_list <- list()
  
  # Loop through each unique sample ID
  for (sample in unique_samples) {
    # Subset data for the current sample
    subset_tumor <- tumor_data[samplesid == sample,]
    subset_plasma <- plasma_data[samplesid == sample,]
    
    # merge
    mg <- merge(subset_tumor, subset_plasma, by=c(1,2,3), suffixes = c("_tumor", "_plasma"))
    
    
    # Create the heatmap plot
    plot <- ggplot(mg, aes(x = status_tumor, y = logR_plasma, col = status_tumor)) +
      geom_boxplot(outlier.shape = NA)+ 
      labs(title =  sample, x = "Status Tumor", y = "logR Plasma") 
    
    # Add plot to the list
    plot_list[[sample]] <- plot
  }
  
  # Arrange the plots into a 3x2 grid
  grid_plot <- do.call(grid.arrange, c(plot_list, ncol = 3, nrow = 2))
  
```
