---
title: "Exploratory analysis of low-pass WGS"
author: 
- Zhan Yinxiu
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  workflowr::wflow_html:
    toc_depth: 4
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo       = FALSE,
                      message    = FALSE,
                      warning    = FALSE,
                      cache      = FALSE,
                      autodep    = TRUE,
                      fig.align  = 'center',
                      fig.width  = 6,
                      fig.height = 6)
```

```{r libraries}
library(caret)
library(ggplot2)
library(stringi)
library(stringr)
library(gridExtra)
library(GenomicRanges)
library(BioCircos)
library(GenomeInfoDb)
library("interacCircos")
```

```{r functions}
mycircos = function(plasma_data, tumor_data, status, sampleid){
  df_plasma = plasma_data[plasma_data$status=="AMP" & plasma_data$sample == sampleid,]
  df_tumor = tumor_data[tumor_data$status=="AMP" & tumor_data$sample == sampleid,]
  
  # Arcs coordinates
  snvChr = df_plasma$chr
  snvStart = df_plasma$start
  snvEnd = df_plasma$end
  # Values associated with each point, used as radial coordinate 
  #   on a scale going to minRadius for the lowest value to maxRadius for the highest value
  snvValues = rep(1, length(snvChr))
  
  # Create CNV track
  tracks = BioCircosCNVTrack(trackname = 'plasma_track', as.character(snvChr), snvStart, snvEnd, snvValues, 
    color = "#CC0000", range = c(0,6), width = 2)
  
  # Arcs coordinates
  snvChr = df_tumor$chr
  snvStart = df_tumor$start
  snvEnd = df_tumor$end
  # Values associated with each point, used as radial coordinate 
  #   on a scale going to minRadius for the lowest value to maxRadius for the highest value
  snvValues = rep(2, length(snvChr))
  # Create CNV track
  tracks = tracks + BioCircosCNVTrack(trackname = 'tumor_track', as.character(snvChr), snvStart, snvEnd, snvValues, 
    color = "blue", range = c(0,6), width = 2)
  
  # Add background
  tracks = tracks + BioCircosBackgroundTrack("arcs_background", colors = "#2222EE")
  
  
  df_plasma = plasma_data[plasma_data$status=="DEL" & plasma_data$sample == sampleid,]
  df_tumor = tumor_data[tumor_data$status=="DEL" & tumor_data$sample == sampleid,]
  
  # Arcs coordinates
  snvChr = df_plasma$chr
  snvStart = df_plasma$start
  snvEnd = df_plasma$end
  # Values associated with each point, used as radial coordinate 
  #   on a scale going to minRadius for the lowest value to maxRadius for the highest value
  snvValues = rep(1, length(snvChr))
  
  # Create CNV track
  tracks = tracks + BioCircosCNVTrack(trackname = 'plasma_track', as.character(snvChr), snvStart, snvEnd, snvValues, 
    color = "pink", range = c(0,6), width = 2)
  
  # Arcs coordinates
  snvChr = df_tumor$chr
  snvStart = df_tumor$start
  snvEnd = df_tumor$end
  # Values associated with each point, used as radial coordinate 
  #   on a scale going to minRadius for the lowest value to maxRadius for the highest value
  snvValues = rep(2, length(snvChr))
  # Create CNV track
  tracks = tracks + BioCircosCNVTrack(trackname = 'tumor_track', as.character(snvChr), snvStart, snvEnd, snvValues, 
    color = "lightblue", range = c(0,6), width = 2)
  
  # Add background
  tracks = tracks + BioCircosBackgroundTrack("arcs_background", colors = "#2222EE")
  
  p = BioCircos(genome = main_chromosomes, tracks, genomeFillColor = "YlOrBr", genomeTicksDisplay = F, genomeLabelDy = 0) 
  return(p)
}
```

# Objectives

In this report, we focus exclusively on comparing copy number variations (CNVs) that are annotated as pathogenic or likely pathogenic, using [ClassifyCNV](https://doi.org/10.1038/s41598-020-76425-3)

```{r data loading}
wgs_tumor_files = list.files("/hpcnfs/scratch/DIMA/zhan/gpelicci/liquid_biopsy_project/data/", pattern = ".tumor.cnv.annotated.tsv", full.names = TRUE) 
wgs_plasma_files = list.files("/hpcnfs/scratch/DIMA/zhan/gpelicci/liquid_biopsy_project/data/", pattern = ".plasma.cnv.annotated.tsv", full.names = TRUE) 

tumor_data = NULL
for(file in wgs_tumor_files){
  tmp = read.delim(file)[,c(2,3,4,5,6)]
  colnames(tmp) = c("chr", "start", "end", "status", "Classification")
  tmp$sample = str_split(basename(file), "\\.")[[1]][1]
  tmp$status[tmp$status=="DUP"] = "AMP"
  tumor_data = rbind(tumor_data, tmp)
}
tumor_data = tumor_data[tumor_data$Classification !="Uncertain significance", ]

plasma_data = NULL
for(file in wgs_plasma_files){
  tmp = read.delim(file)[,c(2,3,4,5,6)]
  colnames(tmp) = c("chr", "start", "end", "status", "Classification")
  tmp$sample = str_split(basename(file), "\\.")[[1]][1]
  tmp$status[tmp$status=="DUP"] = "AMP"
  plasma_data = rbind(plasma_data, tmp)
}
plasma_data = plasma_data[plasma_data$Classification != "Uncertain significance", ]
```

Below, we present the fraction of CNAs detected in tumors that are also found in plasma, and vice versa. On average, around 31\% of CNAs identified in plasma were also detected in tumor samples. Conversely, as expected, a smaller fraction (13\%) of CNAs identified in tumors were also found in plasma. These proportions show minimal variation compared to the overall CNV analysis.

```{r}
results = NULL
percentage = 0.3
for(status in unique(plasma_data$status)){
  for(sampleid in unique(plasma_data$sample)){
    df_plasma = plasma_data[plasma_data$status==status & plasma_data$sample == sampleid,]
    df_tumor = tumor_data[tumor_data$status==status & tumor_data$sample == sampleid,]
    
    gr_plasma <- GRanges(
      seqnames = df_plasma$chr,
      ranges = IRanges(start = df_plasma$start, end = df_plasma$end)
    )
    
    gr_tumor <- GRanges(
      seqnames = df_tumor$chr,
      ranges = IRanges(start = df_tumor$start, end = df_tumor$end)
    )
    
    
    # Find all overlaps
    overlaps <- findOverlaps(gr_plasma, gr_tumor)
    if(length(overlaps)>0){
      # Filter overlaps by percentage
      valid_overlaps <- mapply(function(i, j) {
        overlap_length <- width(pintersect(gr_plasma[i], gr_tumor[j]))
        min_length <- min(width(gr_plasma[i]), width(gr_tumor[j]))
        overlap_length / min_length >= percentage
      }, queryHits(overlaps), subjectHits(overlaps))
      
      # Get indices of valid overlaps
      valid_indices <- overlaps[valid_overlaps]
      
      # Calculate counts
      overlapping_count <- length(unique(queryHits(valid_indices)))
      specific_plasma <- length(gr_plasma) - overlapping_count
      specific_tumor <- length(gr_tumor) - length(unique(subjectHits(valid_indices)))
    }else{
      overlapping_count = 0
      specific_plasma <- length(gr_plasma)
      specific_tumor <- length(gr_tumor)
    }
    tmp = data.frame(fraction = c(overlapping_count/(specific_plasma+overlapping_count), overlapping_count/(specific_tumor+overlapping_count)), 
                     case = c("fraction_relative_plasma", "fraction_relative_tumor"),
                     status = rep(status, 2),
                     sampleid = rep(sampleid, 2)
                     )
    results = rbind(results, tmp)
  }
}
```

```{r}
p1 = ggplot(data = results[results$case == "fraction_relative_plasma" & results$status!="NEUT",], 
       aes(x = as.factor(sampleid), y = fraction, fill = status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(x = "Sample ID", y = "Fraction detected in tumor", fill = "Status", title="CNA in Plasma") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
p2 = ggplot(data = results[results$case == "fraction_relative_tumor" & results$status!="NEUT",], 
       aes(x = as.factor(sampleid), y = fraction, fill = status)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(x = "Sample ID", y = "Fraction detected in plasma", fill = "Status", title="CNA in Tumor") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r, fig.width=12}
grid.arrange(p1, p2, ncol = 2)
```


# Circle plot for visual comparison

Next, we generated a circos plot to visually inspect the overlaps. CNA amplifications from the plasma sample are highlighted in red, while deletions are shown in pink. Similarly, amplifications from the tumor sample are displayed in blue, and deletions in light blue.

```{r}
# Get chromosome sizes for hg38
hg38_sizes <- seqlengths(GenomeInfoDb::Seqinfo(genome = "hg38"))
names(hg38_sizes) = str_remove(names(hg38_sizes), "chr")
main_chromosomes <- as.list(hg38_sizes[names(hg38_sizes) %in% c(as.character(1:22), "X", "Y")])
tumor_data$chr = str_remove(tumor_data$chr, "chr")
plasma_data$chr = str_remove(plasma_data$chr, "chr")
```

## C2 

```{r}
sampleid = "C2"
mycircos(plasma_data, tumor_data, status, sampleid)
```


## E06 

```{r}
sampleid = "E06"
mycircos(plasma_data, tumor_data, status, sampleid)
```



## E23 

```{r}
sampleid = "E23"
mycircos(plasma_data, tumor_data, status, sampleid)
```


## E27

```{r}
sampleid = "E27"
mycircos(plasma_data, tumor_data, status, sampleid)
```


## H7 

```{r}
sampleid = "H7"
mycircos(plasma_data, tumor_data, status, sampleid)
```



## H8

```{r}
sampleid = "H8"
mycircos(plasma_data, tumor_data, status, sampleid)
```





